{\bf Code Generation Recovery Log Replay for In-Memory Database Management Systems}

### 2021-08-03 paper.tex
Updates from Overleaf

% some useful packages
\usepackage{times}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subcaption}

@@

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Code Generation Log Replay for In-memory Database Management Systems}}
\author{Tianlei Pan}
\date{June}
\Year{2021}
\trnumber{}

@@

\keywords{Code Generation, Log Replay, Recovery, Query Compilation}

@@

\begin{abstract}
Instead of interpreting log records, we are extending our query code generation engine to create custom programs to replay log records. We will also be exploring for different optimization techniques to improve the recovery speed of recovery with code generation.
\end{abstract}

@@

\chapter{Introduction}
A \textit{database management system (DBMS)} is a software that is responsible for storing data, analyzing data, and interacting with applications. Depending on the use case, a DBMS can be designed to focus more on either capturing data, or analyzing data. \textit{On-line Analytical Processing (OLAP)} applications focus on reading, analyzing and aggregating cold data that is less likely to be modified. On the other hand, \textit{On-line Transaction Processing (OLTP)} applications work with write-heavy transactions that modify the database frequently.

Consequently, recovery and durability in OLTP applications are more of a pressing issue compared to OLAP systems. Naive or manual backups are usually sufficient for read-only OLAP applications. However, since OLTP databases are being updated by transactions constantly, naive backups will either cause the database to lose significant amount of data, or fail to preserve the \textit{ACID (atomicity, consistency, isolation, and durability)} properties. This requires OLTP applications to develop specialized recovery mechanism to handle these problems. Commercial DBMS often combines Write-ahead Logging (WAL) with checkpoints to preserve database durability.

In addition, OLTP systems also require specialized algorithms and implementations to execute write transactions. Query compilation is an important optimization technique used in OLTP systems to greatly reduce the number of instructions that need to be executed. A DBMS uses a processing model (e.g. Iterator, Vectorization, Materialization) to determine how the DBMS executes a query plan. The processing model is responsible for translating a query into an intermediate expression. The execution engine then breaks down the expression into various tasks for execution. To speed up this process, the DBMS can use an optimization technique known as \textit{Query Compilation}. Query compilation requires the DBMS to generate specialized code for execution tasks to increase execution throughput. These specialized code are often created using an off-shelf compiler into a C/C++ program that implements the functionality of a certain query.

While query compilation has proved successful in increasing the efficiency of a DBMS, it comes with several drawbacks. Implementing a query compilation system can be quite an ordeal. It requires additional knowledge of compiler systems (e.g. LLVM) and huge amounts of engineering work to translate different execution tasks. Moreover, the DBMS also need to modify other components to accommodate with an execution engine reliant on query compilation.

One of such components is the recovery system. Since the execution engine has adapted to executing compiled native code, the DBMS needs to implement an additional execution engine to interpret the log records. Instead, we propose a system design that will extend the query execution engine to create custom programs for replaying log records.

\chapter{Background}
\section{Database Recovery}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Logging.png}
\caption{Write Ahead Logging.}
\label{fig:logging}
\end{figure}
Database recovery systems are responsible for preserving the ACID principles of a DBMS in case of failure. Prior to 1980s, early recovery systems rely heavily on non-volatile memory storage. Most of the DBMSs were disk-based and the recovery protocols assume the existence of dirty pages. System R\cite{system_r}, for instance, creates shadow copy of updated pages and maintains log record updates. Nowadays, both the capacity and reliability of main memory have increased significantly. This greatly improved the capabilities of in-memory DBMSs, but their performance is still be threatened by the need to synchronize non-volatile memory with persistent storage. Modern recovery systems aim to reduce the synchronization overhead.

Recovery systems consist of two components: logging and recovery. Logging is the action of storing transactional information on disk. These information are stored in a special data structure called a log record. A log record can either contain physical information (e.g. changes made to a specific physical address) or higher-level information (e.g. user-input query). In Write-ahead logging \cite{aries}, the DBMS records database changes in a log file before the changes can be persisted on disk. Each log record in the log file usually consists of redo records (install the effects of a committed transaction) and undo records (remove the effects on an incomplete or aborted transaction). Since in-memory DBMSs do not persist uncommitted changes, undo records are no longer required for the recovery system.

Recovery is the action of restoring the database system into a consistent state. If available, the DBMS first attempts to restore its previously recorded checkpoint. The database then initiates the log replay process to go through every log record stored since that checkpoint. SiloR\cite{silo_r} uses physical logging, checkpoints and an epoch-based recovery system. Transactions are assigned with different epochs and can only be persisted when their epochs are smaller then the current persistent epoch. VoltDB\cite{malviya14} uses a form of physiological logging that reduces transaction execution overhead. SAP HANA\cite{lee18}, however, uses a log-less approach through parallel table replication.

\section{Query Compilation / Code Generation}
When a query arrives, the DBMS parses the query into an expression tree. Eventually, the optimizer of the DBMS converts the expression tree into a physical plan tree. The physical plan represents how the query will be executed. It usually consists of operators that specify physical operations on the DBMS (e.g. inserting values into a certain memory location). For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child. The execution engine then uses the plan tree for execution.

The Iterator model\cite{volcano} is widely used in many disk-based DBMS (e.g. ORACLE RDBMS, MySQL, Microsft SQL Server) to interpret plan trees. The model "iterates" over each operator by implementing a \dbSQL{next()} function in each operator that returns the next tuples to be operated. The expressions inside the operators are evaluated using the provided tuples and generates output tuples. This approach suffers from frequent virtual function calls for in-memory DBMS. The Materialization model reduces this overhead by processing all the input tuples and emitting all the outputs at once ("materializing" the result as a single row or column). The Materialization model is suitable for OLTP DBMS since each OLTP query only accesses a small number of tuples at a time. However, since the Materialization results are stored in an in-memory buffer, the execution throughput is then bound by the size of the available memory space. While disk-based DBMSs are less concerned about the in-memory access since most of their operations are disk-based, this is especially critical for in-memory DBMSs.

Vectorization and Query Compilation are two state-of-the-art techniques employed for in-memory DBMSs to reduce memory overhead\cite{kersten18}. The Vectorization model is a variant of the Iterator model. Instead of emitting a single tuple at a time, the Vectorization model emits a batch of tuples at a time to amortize the cost of memory access. The functions that operate on these tuples are also specialized to be able to take in a batch of tuples at a time and more efficient compared to iterating over a single tuple at a time.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/QueryCompilation.png}
\caption{Query Compilation architecture.}
\label{fig:query_compilation}
\end{figure}

Query Compilation eliminates the cost of interpretation by compiling the physical plan to machine code. The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls. Before being compiled to machine code, the physical plan is often first converted into an intermediate representation (e.g. C/C++ code, LLVM IR, or custom Domain Specific Language) that follows the grammar of an imperative language. This simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resembles closely to SQL statements.

Hyper\cite{neumann11} is one of the pioneers of query compilation. It uses a data-centric compilation approach that splits the query plan into pipelines and compiles each pipeline separately. Vectorwise\cite{raducanu13} chooses to pre-compile huge amounts of micro function primitives that perform simple operations on typed data. During runtime, the DBMS can directly invoke these compiled primitives for query execution.

\chapter{Method}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Proposal.png}
\caption{Code Generation Recovery Mechanism.}
\label{fig:codegen_recovery}
\end{figure}

Our proposal is to extend the recovery layer and query compilation layer to support the interpretation of log records. This approach can be realized in three steps:

\begin{enumerate}
\item Extract information from log records.
\item Convert the information into expressions that can be accepted and compiled by the code generation engine (Code Generation Step).
\item Run the compiled machine code using the execution engine.
\end{enumerate}

This code generation approach brings around several benefits. Firstly, it significantly reduces engineering overhead. Compiling contents of the log records into machine code allows the recovery manager to seamlessly integrate with the built-in execution engine. This removes the need to specifically construct code to handle tasks that are not compiled to machine code.

\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
 \caption{Index Update During Recovery.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RecoverConstraint.png}
 \caption{Constraint Checking During Recovery.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{Problems faced by conventional recovery in code generation DBMS.}
\label{fig:log_record}
\end{figure}

Secondly, it also resolves tricky problems that will appear when writing specialized code for log records. An example of this is index updates. Without code generation, the recovery manager needs to specifically reconstruct the index given an insert or update record. With code generation, the index updates are taken care directly by the execution engine that already contains high-performance specialized code for index maintenance. Similar problems, such as constraint checking

Our proposed architecture is implemented and tested in \textit{Noisepage} \cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University. Noisepage is built in C++ and uses the PostgreSQL wire protocol for user communication. The DBMS supports ACID transactions and \texttt{SNAPSHOT-ISOLATION}. Noisepage uses the conventional WAL recovery scheme with physical logging and depends on query compilation for query execution. We will dive into the details of Noisepage's recovery structure in the following sections.

\section{Recovery System Architecture}
\subsection{Transaction}
Transactions are the atomic operations of any DBMS. Noisepage uses a transaction to perform physical updates on the database and physical logging to disk. Noisepage uses a multi-versioned delta store transaction engine that ensures \texttt{SNAPSHOT-ISOLATION}. The transaction engine allows non-blocking reads over writes and vice versa, but does not allow write-write conflicts on a per tuple basis. A transaction's lifecycle begins when the transaction begins, and ends after the log manager has serialized its changes to disk.

Each data tuple in noisepage is uniquely identified through a tuple slot that stores the offset of a tuple. When the DBMS inserts a tuple into a table, it creates a new tuple slot that points to a memory location within the table. Subsequent updates to the tuple does not create direct copies of the tuple, but stores delta information (redo records) about the updates instead. These redo records are essential for executing log replay correctly. The structure of a redo record is shown later in \ref{fig:log_record}.

A transaction uses a buffer (redo buffer) that stores all the delta records that will appear in its lifetime. Each redo buffer has a fixed size of 4096 bytes and is allocated from a centralized buffer pool. Each time a transaction needs to change the contents of the DBMS, it will attempt to request space from the redo buffer to dump a corresponding delta record of the change. If the redo buffer runs out of space, the transaction will replace the current buffer with a new buffer from the buffer pool. The redo buffer allows the logging components to process the changes before the transaction is completed.

When a transaction commits, the transaction engine creates a commit record that contains a timestamp that refers to the oldest active transaction from the timestamp manager. The commit record is appended to the redo buffer. The redo buffer is then carried over to the log manager, where the changes will be serialized to disk. When a transaction aborts, the transaction creates an abort record that will prevent corresponding records to be replayed during recovery. Abort records are essential; since the redo buffer is persisted to disk by the log manager once it is full, it is possible for aborted transactions to persist on disk. Aborted transactions that are persisted on disk are rare, however, since it would be difficult for a typical OLTP transaction to fill up the entire redo buffer.

\subsection{Write-Ahead Logging}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/WAL.png}
\caption{Logging Architecture. Log manager uses the log serializer and log consumers to convert redo buffers into log records.}
\label{fig:wal}
\end{figure}

The log manager is responsible for performing write-ahead logging in Noisepage. All the changes are serialized to disk as logs before they are performed on the DBMS storage. Each log record is self-contained and does not require any additional metadata to be replayed. The log manager consists of two separate tasks that live in different threads: the Log Serializer tasks and the Log Consumer tasks.

\subsubsection{Log Serializer}
The log serializer task receives redo buffers from a transaction once it is full, or when the transaction decides to commit. The log serializer breaks down the redo buffers into raw memory segmented into fixed-size buffers of 4096 bytes. These buffers will be consumed by the Disk Log Consumer for persistence.

The log records must be serialized in the correct order for recovery to perform correctly. Different transactions can share a same redo buffer that is filled with each of their own log records. Moreover, it is possible for transactions to appear in a non-serial order relative to their begin timestamp. However, the log records performed by an individual transaction is guaranteed to appear in the same order as they were created. This property helps ensure \texttt{SNAPSHOT-ISOLATION} property of Noisepage during log replay.
\subsubsection{Log Consumer}
Buffers supplied by the log serializer can be consumed by different log consumer tasks. Each log consumer is given a copy of the original buffer to achieve parallelism. Noisepage uses two type of log consumers: disk log consumer and replication log consumer.

The disk log consumer repeatedly polls for new buffers from the log serializer and writes the changes to disk. To improve performance, disk log consumer uses group commit that allows a batch of changes to be committed over a time period specified by the user. Under default settings, the disk log consumer writes down changes every 10 microseconds of if more than 1 MB of data has been written since the last write.

The replication log consumer also polls for new buffers, but instead of persisting them to disk, sends the new buffers over across the network to database replicas.

\subsection{Log Replay}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/RecoveryManager.png}
\caption{How recovery operates in Noisepage.}
\label{fig:recoverymanager}
\end{figure}

Noisepage uses four different record types: \textit{Redo, Delete, Commit, Abort}. Redo and Delete records are not processed immediately. Instead, they are placed inside a deferred changes buffer, ordered by their transaction timestamp. An Abort record removes the corresponding Redo/Delete record from the buffered changes. A Commit record initiates a transaction to iterate over the deferred changes buffer. Once all the deferred changes are processed, the buffer is cleared and the transaction is committed. The transaction aborts if any log record is malformed or any recovery operation fails.

\subsubsection{Data Structure}
\begin{figure}[H]
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/RedoRecord.png}
 \caption{Redo Record Structure.}
  \label{fig:pipeline_graph}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=0.9\textwidth]{images/DeleteRecord.png}
 \caption{Delete Record Structure.}
  \label{fig:pipeline_code}
\end{subfigure}
\caption{The internal log structures.}
\label{fig:log_record}
\end{figure}

A physical log record consists of a header followed by a record body. The header stores metadata about the log record itself, including its record type, size and begin timestamp of the transaction that generated the log record as mentioned in Section 3.1.1. The record body contains physical information that is necessary for recovery and varies depending on the record type. Depending on the value type, the values may bee stored in-line or not in-inline. Simple types such as \texttt{INTEGER}, \texttt{FLOAT} can be stored in-line, while \texttt{VARLEN} entry may stores the pointer in the values column that points to another location in the record.

Tuple slots in the log records will no longer be valid memory locations during recovery. Instead, the recovery manager creates an internal mapping from original tuple slots to a new tuple slots that will be retrieved after new values are inserted. The mapping allows the recovery manager to apply changes to the correct memory locations in the new memory environment.

Since only Redo and Delete records will effectively modify physical storage, we only need to analyze the record body of those two record types. Modifications of a query to the DBMS can be resolved down to the modification of tuple slots on the physical level.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{images/Delta.png}
\caption{Redo Record Delta.}
\label{fig:logging}
\end{figure}

 While storing tuple slots is sufficient for delete records, Redo Records require additional data to perform Recovery Insert and Update operations. These data are kept in the Record Delta, which stores a mapping from column ids to values.

\subsubsection{Recovery Operation}
A Noisepage recovery step has three different operations:
\begin{itemize}
    \item \textbf{Insert}: Represented by a Redo Record. The recovery manager inserts the values stored within the redo record into the given tuple slot and updates the indexes.
    \item \textbf{Delete}: Represented by a Delete Record. The recovery manager deletes the tuple slot specified by the delete record and updates the indexes.
    \item \textbf{Update}: Represented by a Redo Record. The recovery manager updates the tuple slot using the values within the redo record and updates the indexes.
\end{itemize}

While both Insert and Update are represented using the same Redo Record structure, they can be differentiated by keeping track of all the tuple slots found in log records. If a tuple slot inside a Redo Record is unseen before, then the Redo Record represents an Insert operation. Otherwise, the Redo Record represents an Update operation.

\section{Recovery Code Generation}
The code generation structure in Noisepage follows from the data-centric compilation approach in \cite{neumann11}, as seen in \ref{fig:query_compilation}. A query is interpreted, optimized and converted into a physical plan tree. A plan tree consists of plan nodes and specifies how the query should be executed on the physical level of the database.

A physical plan tree/node is then converted into an operator pipeline through a plan translator. Each plan translator correspond to a plan node in the original plan tree. A pipeline contains an ordered collection of relational operators that are not yet materialized, ended with a pipeline breaker (e.g. sort tuples, creating a hash table). The pipeline is only materialized whenever a pipeline breaker has been encountered.

>>>>

% some useful packages
\usepackage{libertine}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{amsmath, lipsum, afterpage}
\usepackage[numbers,sort]{natbib}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered,
%pdfborder=0 0 0,  %removes outlines around hyper links in online display
]{hyperref}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{caption}

@@

\hypersetup{colorlinks,linkcolor={blue},citecolor={blue},urlcolor={red}}

@@

\title{ %% {\it \huge Thesis Proposal}\\
{\bf Code Generation Recovery Log Replay for In-Memory Database Management Systems}}
\author{Tianlei Pan}
\date{June}
\Year{2021}
\trnumber{}

@@

\keywords{Code Generation, Log Replay, Recovery, query compilation}

@@

\begin{abstract}
Code generation is an optimization technique for improving query execution throughput by compiling query plans into native code. This technique, however, leads to design challenges for the recovery system of a database management system. The log replay process will be disjoint from the built-in execution engine that has adapted to operate efficiently on compiled code. This leads to the implementation of a separate execution engine to deal with the execution of log records, which can be a huge waste of engineering efforts. To resolve this design conflict between code generation and database recovery, we present a unified approach to support both query execution and log replay in a code-generation-based DBMS. We ask the recovery system to convert log records into compiled code that will be easily accepted by the execution engine. Our results show that while our approach incurs a higher performance overhead compared to using a separate execution engine, it requires much less engineering effort and is potentially superior in aspects such as index updates and constraint checking.
\end{abstract}

@@

\chapter{Introduction}
Database management system (DBMS) is a category of software that are responsible for storing data, analyzing data, and interacting with applications. In 1970, E.F. Codd proposed the relational model of data,

Depending on the use case, a DBMS can focus more on either capturing data or analyzing data. On-line Analytical Processing (OLAP) DBMSs focus on reading, analyzing and aggregating data that is less likely to be modified. On the other hand, On-line Transaction Processing (OLTP) DBMSs support write-heavy transactions that modify the database frequently~\cite{conn05}.

Regardless of the type of a DBMS, both OLTP and OLAP DBMSs are susceptible to failures. The DBMS may shutdown unexpectedly, fail to execute a query, or cease to function due to corrupted data. These failures threaten the integrity of the database and make the DBMS unreliable. Therefore, it is important for a DBMS to distinguish between different kinds of failures in order to develop mechanisms to maintain integrity.

A DBMS can encounter three types of failures: Transaction Failures, System Failures, and Media Failures\cite{theo83}. Transaction Failures are the most common type of DBMS failures. Transaction Failures occur when a transaction fails to commit, either at its own request (e.g., logical errors) or on behalf of the DBMS (e.g., resource unavailability). When a transaction fails, the system needs to react within its lifetime to prevent data inconsistency~\cite{aries}. The DBMS needs to decide whether and how it should undo the changes of the failing transaction. Since a DBMS is running transactions rapidly while its active, Transaction Failures can occur as much as 100 times per minute\cite{theo83}. System Failures occur due to hardware failures (e.g., power outage), operating system faults (e.g., insufficient memory) or DBMS exceptions. Each of these trigger events can cause the DBMS to shutdown unexpectedly and uncontrollably. System Failures occur less frequently compared to Transaction Failures, but take a much longer time for the DBMS to recover. The database needs to pinpoint time of failure and what changes the DBMS failed to record. Media Failures~\cite{sauer17} happen when the underlying storage device of a DBMS fails (e.g., disk head crash, bad sectors). This type of failure causes permanent data loss to the DBMS. The only way for a DBMS to recover from Media Failures is to restore from a secondary backup storage.

To handle different types of failures, a DBMS needs to develop a recovery system that will restart the database correctly in case of failure. A widely adopted recovery system is to combine write-ahead logging during transaction execution with crash recovery that uses redo (install changes to the database) and undo (remove changes from the database) processes, represented by ARIES\cite{malviya14}. The ARIES (Algorithm for Recovery and Isolation Exploiting Semantics)~\cite{aries} protocol is a recovery method developed by IBM in the 1990s. It guarantees database integrity in the fact of Transaction, System and Media Failures~\cite{aries}. In ARIES, the DBMS records transaction modifications in durable log records before the DBMS propagates changes to a database page to disk. During crash recovery, the DBMS applies changes from the log records to the database with redo and undo processes (i.e., log replay) ~\cite{franklin97}.

However, the original ARIES paper focused on disk-based DBMSs. Recently, we have seen a rapidly increasing number of in-memory DBMSs due to technological advancement in semiconductor memory~\cite{gar92}. Many in-memory DBMSs avoid strictly adhering to the ARIES protocol for recovery. This is due to the fact that a lot of the concepts in ARIES no longer apply to in-memory DBMSs. For instance, ARIES uses undo records to revert changes that have been applied to the database. However, undo records are excessive for many in-memory DBMS recovery systems~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}. Performance concerns is another reason why in-memory DBMSs stray from the ARIES protocol. Logging requires the DBMS to interact with the disk. Therefore, logging I/O is a major bottleneck for an in-memory DBMS. This prompts in-memory DBMSs to minimize logging traffic~\cite{faeber17}. An example is the recovery of database indexes~\cite{faeber17}. Disk-based DBMSs log updates to index structures (e.g., B+ Tree) that allow faster recovery during log replay~\cite{aries}. In comparison, in-memory DBMSs do not log index updates~\cite{hekaton2013, noisepage, malviya14, silo_r} and choose to reconstruct indexes from scratch during log replay. Despite the differences, many in-memory DBMSs still use some form of logging~\cite{noisepage, malviya14, hekaton2013, silo_r} in their recovery systems.

In addition to changes in recovery systems, in-memory DBMSs adopt various optimization algorithms for their execution engines\cite{kersten18}. These optimizations aim to increase execution throughput by reducing either the number of instructions that a DBMS executes to run a query, or the clock cycles per instruction (CPI)~\cite{boncz05}. Query compilation/code generation is an important optimization technique used by DBMSs to greatly reduce the number of instructions that the CPU needs to execute\cite{kersten18}. During execution, the DBMS breaks down a query into various tasks~\cite{noisepage, neumann11}. To speed up the execution of those tasks, the DBMS can compile them into native code (e.g., C/C++) with an off-shelf compiler\cite{boncz05}. Query compilation leads to faster query execution because it specializes both the data structures (e.g., hash table) and access methods of a DBMS towards execution efficiency~\cite{neumann11}. Moreover, the compiled code can be optimized around locality that increases the chance of a data tuple being propagated between operators in CPU registers\cite{menon17}.

While query compilation succeeds in increasing the execution throughput of a DBMS, it comes with several drawbacks. Implementing a query compilation system requires additional knowledge of compiler systems (e.g., LLVM) and huge amounts of engineering work to translate different execution tasks\cite{kersten18}. The DBMS also generates low-level machine code that is hard to understand and debug~\cite{kersten18}.

Moreover, the DBMS recovery system is isolated from the query compilation execution engine (i.e., a component of a DBMS that is responsible for execution). Most of the query compilation DBMSs we have surveyed~\cite{noisepage, neumann11, paroski16, craig14} have no conduits between their recovery systems and execution pipelines. The recovery system implements its own functionalities to update the database, but some of these functionalities already exist in the execution engine. For instance, we mentioned that in-memory DBMSs rebuild the indexes during log replay. When a redo process install changes on a table, it also retrieves the indexes on that table and update them (index update). However, the execution engine has already implemented this index update functionality. As a result, the DBMS contains two implementations that provide identical functionalities. Similar reasoning applies to other functionalities (e.g., constraint checking, trigger events) that may be used during log replay. The number of those functionalities in the recovery system will eventually amount to the scale of a separate execution engine that is built specifically for log replay.

Therefore, if we can find a way to unify a DBMS's recovery system and execution engine, we can drastically reduce engineering overhead. Furthermore, we believe this unification also increases replay efficiency. Unless the rewritten functionalities in the recovery manager are executed as native machine code, they will be less efficient compared to those in the execution engine that uses query compilation. Therefore, if the DBMS allows the recovery system to invoke functionalities within the execution engine, the log replay process should be more efficient.

\section{Contribution}
We present a system design that will extend the execution engine to create custom programs for replaying physical log records. Our design converts log records into a format that the DBMS can compile into native code to be accepted by the execution engine. We will show that our approach removes the need for introducing extra implementations into the recovery system for functionalities that already exist in the execution engine. We will also verify our assumption that by utilizing the code generation execution engine, the DBMS will have increased log replay efficiency.

We implement our approach in NoisePage\cite{noisepage}, a self-driving in-memory DBMS developed at Carnegie Mellon University. NoisePage is built in C++ and uses the PostgreSQL wire protocol for user communication. The DBMS also depends on query compilation for its execution engine. Our experimental results show that while the code generation recovery approach falls short of recovery throughput compared to baseline recovery, it is more efficient in certain areas (e.g., index maintenance, table constraints) and has higher scalabiltiy for a table with a larger number of columns.

We structure the remainder of the thesis as follows. In Chapter 2, we provide more background information on database recovery, code generation, and motivations behind our approach. In Chapter 3, we present the recovery and code generation architecture in NoisePage and how we implemented our approach in this DBMS. In Chapter 4, we evaluate our implementation by comparing it against NoisePage's builtin recovery system. We discuss related works in Chapter 5 and conclude our thesis in Chapter 6.

\chapter{Background}
In this chapter, we discuss more background information on database recovery and code generation. We then explain how the two components interact within a DBMS and what changes need to be made for them to cooperate.

\section{Transactions}
Transactions are the atomic operations of any DBMSs~\cite{theo83}. They appear throughout a DBMS, including the recovery system and the execution engine responsible for code generation. A transaction is a sequence of interactions with the database. These interactions can be retrieving data, inserting data, removing data, etc.

\section{Database Recovery}
A DBMS ensures database integrity by satisfying the following conditions:
\begin{itemize}
    \item \textbf{Durability of Updates}~\cite{agrawal89}: All the changes made by committed transactions are durable.
    \item \textbf{Failure Atomicity}~\cite{franklin97}: None of the changes made by aborted or failed transactions are persisted visible after recovery.
\end{itemize}

Many DBMSs~\cite{noisepage, malviya14, hekaton2013, silo_r, azure, postgres, mysql} combine logging and recovery protocols to preserve these two conditions in case of failure. Logging is the action of storing information about committed transactions on disk, while recovery is the action of restoring the database system into a consistent state.

The DBMS stores logging information in a special data structure called a log record. A log record contains physical changes (e.g., changes made to a specific physical address) performed on the DBMS, or higher-level information (e.g., user-input query). In write-ahead logging \cite{aries}, the DBMS records physical database changes in a log file before the DBMS persists the changes on disk. A transaction is not allowed to commit until the DBMS persists its corresponding log record.

For disk-based DBMSs, a log record in the log file can be either a redo record or an undo record~\cite{aries}. During recovery, a redo record installs the effects of committed transactions, while an undo record removes the effects on incomplete or aborted transactions. In-memory DBMSs, on the other hand, do not persist uncommitted changes on disk~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}. This is because disk-based DBMSs generate dirty pages. These are pieces of modified entries in the database that reside in the memory, but have not yet been persisted to disk by the DBMS. However, in-memory DBMSs do not write dirty data to persistent storage~\cite{noisepage, malviya14, hekaton2013, scuba14, silo_r}. Therefore, in-memory DBMSs do not generate dirty pages, nor do they need to keep undo records. As a result, undo records are no longer required for their recovery systems. Moreover, in-memory DBMSs do not need to generate log records for indexes~\cite{faeber17}. During recovery, the DBMS reconstructs the database with redo records reconstructs and create the indexes simultaneously~\cite{malviya14, hekaton2013}. For each redo record, the recovery system finds the table that it modified, then retrieves all of its indexes from the database catalog ((\cref{fig:recovery_index_update}). Finally, for each index, the recovery system either updates it with new values, or remove entries from the index.

\begin{figure}[t!]
\centering
\centering
\includegraphics[width=0.9\textwidth]{images/RecoverIndex.png}
\caption{Index Update During Recovery.}
 \label{fig:recovery_index_update}
\end{figure}

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/RecoveryCheckpoint.png}
\caption{Recovery with Log Records and Checkpoints.}
\label{fig:recovery_checkpoint}
\end{figure}

Aside from logs, recovery systems can store complete snapshots (i.e., checkpoints) of the database. The DBMS can decide to take checkpoints based on a certain rule (e.g., timed-interval, number of transactions). Checkpoints reduce the time required for the database to recover. The DBMS can directly revert to a previous snapshot without replaying any log records before that snapshot. DBMSs often store both log records and checkpoints. During recovery, the DBMS finds its most recent snapshot, then replays any log records that are stored after that snapshot (\cref{fig:recovery_checkpoint}).

\section{Query Compilation / Code Generation}
\begin{figure}[t!]
\centering
\includegraphics[width=0.5\textwidth]{images/PhysicalPlanExample.png}
\caption{Physical Plan that corresponds to \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100}.}
\label{fig:physical_plan_example}
\end{figure}
When a query arrives, the DBMS parses the query into an abstract syntax tree (AST). The optimizer of the DBMS then converts the AST into a physical plan tree. The physical plan represents how the DBMS will execute the query. It consists of operators that specify physical operations on the DBMS (e.g., inserting values into a certain memory location). For instance, \dbSQL{UPDATE x SET col1 = -col1 WHERE col1 BETWEEN 0 AND 100} can be represented by an update plan tree with a sequential plan as a child (\cref{fig:physical_plan_example}). The execution engine then uses the plan tree for execution.

In a DBMS that uses query interpretation, its execution engine (i.e., a component of a DBMS that is responsible for execution) processes the physical plan by traversing the plan tree. Therefore, for every query, DBMS needs to follow pointers and resolve branching conditions (e.g., if, switch statements). This incurs overhead caused from virtual function calls, branch mispredictions and instruction cache misses.

\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/QueryCompilation.png}
\caption{Query Compilation Architecture.}
\label{fig:query_compilation}
\end{figure}

Query compilation eliminates the cost of query interpretation by compiling the physical plan to machine code that is specific for that query. The machine code of the corresponding physical plan can then be executed repeatedly by the execution engine, removing any need for virtual function calls and branching resolutions. Generally, there are two ways of compiling queries. In the first approach, the DBMS creates source code that is compiled into native code with an external compiler (e.g., gcc) ~\cite{krikellas10, klonatos14}. This approach is used in Amazon Redshift and pre-2016 MemSQL~\cite{paroski16}. In the second approach, the DBMS first converts physical plan into an intermediate representation (e.g., LLVM) that follows the grammar of an imperative language. This approach simplifies the process of converting the physical plan into machine code, as the intermediate representation form is designed to resemble SQL statements. Moreover, it removes the need for the DBMS to use the compiler as an external process. This approach is adopted by NoisePage~\cite{menon17}, Hyper~\cite{neumann11}, Hekaton~\cite{craig14}, and SingleStore~\cite{paroski16}.

\section{Motivation}
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/Motivation.png}
\caption{Possible ways to integrate a physical logging system with code generation.}
\label{fig:motivation}
\end{figure}
While code generation is a powerful optimization technique, it poses design challenges for the recovery system. Both the execution engine and physical log records perform identical physical operations on a database table. The native code generated through query compilation operates directly on physical tables. For physical log records, they are already designed as a data structure that represents physical changes made to the database. During recovery, the DBMS simply replays those changes on the table.

However, the recovery system has no means of reusing functions within the execution engine to install physical changes from log records. The execution engine does not accept physical log records because it follows along the query compilation pipeline. This pipeline transforms one logical representation of a query to another, until the query becomes native code. There is no place for physical log records to exist within this transformation pipeline.

There are two solutions to this problem. One solution is to re-implement functions to install physical changes on data tables within the recovery system. However, this approach completely isolates the recovery system from the execution engine. This means that if the recovery system wishes to use a functionality within the execution engine, it has to rewrite that same functionality within the recovery system (\cref{fig:problem}).

The other solution is to integrate physical logging with the execution engine. \cref{fig:motivation} shows four possible paths that allow the execution engine to accept physical log records. Converting the log record directly into native code requires the implementation of a new compiling framework, which makes this option infeasible. Therefore, the closest point where the log records can reach before entering the execution engine is physical plans. We then reach the conclusion to convert log records into physical plans. Firstly, compared to raw queries or abstract syntax trees, physical plans are closer to the end of the query compilation pipeline. If we convert a log record into an abstract syntax tree (AST), the DBMS still needs to do additional work to convert the AST into a physical plan. Secondly, we will show in later sections that it is enough for physical plans to represent contents of a log record.

If the DBMS can convert log records into physical plans, the recovery system can rely on code generation for recovery operation. This code generation recovery approach brings around several benefits. Firstly, it significantly reduces engineering overhead. Compiling contents of the log records into machine code allows the recovery system to seamlessly integrate with the built-in execution engine. We have explained in Section 2.1 that the recovery system needs to perform extra operations (e.g. index updates, constraint checking in \cref{fig:problem}) besides updates to the data tables. These functionalities exist within the execution engine. If we allow the recovery system to reuse existing functionalities by communicating with the execution engine through native code, then the recovery system no longer needs to implement its own functionalities. Secondly, utilizing the code generation execution engine provides more efficient functionalities. A function implemented by the recovery system will not be as efficient as one that exists in the execution engine.

These potential benefits motivate us to design a code generation recovery system based on conversion from log records to physical plans. In the next chapter, we will present the implementation of our approachhbmmn .

\chapter{Method}
\afterpage{%
\begin{figure}[t!]
\centering
\includegraphics[width=0.9\textwidth]{images/Method.png}
\caption{Code Generation Recovery Mechanism.}
\label{fig:codegen_recovery}
\end{figure}}
